Unsupervised embeddings capture latent structure of scientific mobility

### 2020-02-24 paper/current/main.tex
revise gravity part and add mathmetical proof

% Math
\usepackage{amsmath}
\usepackage{amssymb}
\DeclareMathOperator*{\argmin}{\arg\!\min}
\DeclareMathOperator*{\argmax}{\arg\!\max}

@@

% for quick author comments 
\usepackage[usenames,dvipsnames,svgnames,table]{xcolor}
\definecolor{light-gray}{gray}{0.8}
\def\del#1{ {\color{light-gray}{#1}} }
\def\yy#1{ {\color{red}\textbf{yy: #1}} }
\def\dk#1{ {\color{red}\textbf{dk: #1}} }

@@

As a result of our disambiguation procedure, some affiliations are mapped to two organizations, one specific, and one more general.
For example, any author affiliated with the "Indiana University Bloomington" will also be listed as being affiliated with the "Indiana University System", a more general designation for all public universities in Indiana. 
However, a more general organization may not always occur alongside the more specific one.
For example, a researcher affiliated with the smaller regional school "Indiana University South Bend" will be listed as affiliated with only the "Indiana University System". 
We identify all specific organizations that always co-occur along a more general one.
For every career trajectory that includes one of these specific organizations, we remove all occurrences of the more general organization;
trajectories containing only a general designation are not altered.

@@

We also calculate two forms of proximity between organizations. 
The geographic distance, $g_{ij}$, is the pairwise geographic distance between organizations.
Geographic distance is calculated as the great circle distance, in kilometers, between pairs of organizations. 
When less than one kilometer, a distance of one is imputed. 
The embedding distance, $d_{ij}$, is calculated as the pairwise cosine distance between vectors from the learned organization embedding, where $v_{i}$ and $v_{j}$ represent the vectors and $d_{ij} = \frac{v_{i} \dot v_{j}}{\lVert v_{i} \rVert \lVert v_{j} \rVert}$.

@@

\begin{equation}
	\label{eq:gravity_basic}
	f_{ij} = C\frac{m_{i}m_{j}}{f(r_{ij})}
\end{equation}

where $m_{i}$ and $m_{j}$ are the yearly-average size of each organization during the time window, and $f(r_{ij}$ is a function of the distance between the two organizations, chosen as to fit the data, and $C$ is a constant. 
Form this standard formulation we also define what we term the \textit{flux}, or the expected flow between organizations given their size, as a function of distance,

\begin{equation}
	\label{eq:gravity_flux}
	F_{ij} = \frac{f_{ij}}{m_{i}m_{j}} = Cf(r_{ij})
\end{equation}

\dk{We need to mention why we choose these distance functions}

We consider separate variants of $f(r_{ij})$ for geographic distance, $g_{ij}$, and the embedding distance, $d_{ij}$. 
For geographic distance, we use the power-law function of the gravity law, with $f(g_{ij})= g^{\alpha}_{ij}$, where $\alpha$ is estimated from the data. 
For embedding distance , we use the exponential function, with $f(d_{ij}) = e^{\beta d_{ij}}$, where $\beta$ is estimated from the data. 
To estimate parameters $\alpha$ and $\beta$ are estimated by log-transforming each variant of \label{eq:gravity_flux}, obtaining Eq.~\ref{eq:linear_flux_geo} and Eq.~\ref{eq:linear_flux_emb} in the form of linear equations, and fitting parameters using linear regression.

\begin{equation}
	\label{eq:linear_flux_geo}
	\ln(F_{ij}) = \ln(C) - \alpha \ln(g_{ij})
\end{equation}

\begin{equation}
	\label{eq:linear_flux_emb}
	\ln(F_{ij}) = \ln(C) + \beta d_{ij}
\end{equation}

@@

Let $f_{ij}$ represents the total flux (the number of people who moved) between $i$ and $j$. 
Because we are ignoring the directions, $f_{ij} = f_{ji}$. 
The total flux from or to $i$ can be written as $\mu_{i} = \sum_k f_{ik}$. 
Then the conditional probability $p(j|i)$ is simply
\begin{equation}\label{fig:flux_conditional_prob}
p(j|i) = \frac{f_{ij}}{\mu_i}. 
\end{equation}

Let us assume that we have identified the best embedding vector $v_i$ for every institution. 
The skip-gram model maximizes the average log probability
\begin{equation}
\frac{1}{T}\sum_{t=1}^{T} \sum_{-c \le j \le c, j \neq 0} \log p(w_{t+j}|w_t),  
\end{equation}
where 
\begin{equation}
p(w_O|w_I) = \frac{\exp(v'_{w_{O} }\cdot v_{w_{I}})}{\sum_{w=1}^{W} \exp(v'_w \cdot v_{w_{)}}}, 
\end{equation}
where $v_w$ and $v'_w$ are `input' and `output' vector representations of $w$.

The best embedding should be able to closely approximate the true conditional probability Eq.~\ref{fig:flux_conditional_prob}.  \yy{why? can we prove this?}

\begin{equation}
p(j|i) = \frac{f_{ij}}{\mu_i} \simeq \frac{\exp(v'_j \cdot v_i)}{\sum_{w=1}^{W} \exp(v'_w \cdot v_i)}
\end{equation}

>>>>

% Math
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{newtxmath}
\DeclareMathAlphabet{\mathpzc}{T1}{pzc}{m}{it}
\DeclareMathOperator*{\argmin}{\arg\!\min}
\DeclareMathOperator*{\argmax}{\arg\!\max}

@@

% for quick author comments 
\usepackage[usenames,dvipsnames,svgnames,table]{xcolor}
\definecolor{light-gray}{gray}{0.8}
\def\del#1{ {\color{light-gray}{#1}} }
\def\yy#1{ {\color{red}\textbf{yy: #1}} }
\def\dk#1{ {\color{red}\textbf{dk: #1}} }
\def\js#1{ {\color{red}\textbf{js: #1}} }

@@

As a result of our disambiguation procedure, some affiliations are mapped to two organizations, one specific, and one more general.
For example, any author affiliated with the "Indiana University Bloomington" will also be listed as being affiliated with the "Indiana University System", a more general designation for all public universities in Indiana. 
However, a more general organization may not always occur alongside the more specific one.
For example, a researcher affiliated with the smaller regional school "Indiana University South Bend" will be listed as affiliated with only the "Indiana University System". 
We identify all specific organizations that always co-occur along a more general one.
For every career trajectory that includes one of these specific organizations, we remove all occurrences of the more general organization;
trajectories containing only a general designation are not alted.

@@

We also calculate two forms of proximity between organizations. 
The geographic distance, $g_{ij}$, is the pairwise geographic distance between organizations.
Geographic distance is calculated as the great circle distance, in kilometers, between pairs of organizations. 
When less than one kilometer, a distance of one is imputed. 
The embedding distance, $d_{ij}$, is calculated as the pairwise cosine distance between vectors from the learned organization embedding, where $v_{i}$ and $v_{j}$ represent the vectors and $d_{ij} = 1 - \frac{v_{i} \dot v_{j}}{\lVert v_{i} \rVert \lVert v_{j} \rVert}$.
\js{cosine similarity itself is not the distance}

@@

\begin{equation}
	\label{eq:gravity_basic}
	f_{ij} = C*m_{i}m_{j}*f(r_{ij})
\end{equation}

where $m_{i}$ and $m_{j}$ are the yearly-average size of each organization during the time window, and $f(r_{ij})$ is a distance-decaying function  between the two organizations. There are two popular forms of the $f(r_{ij})$. The first one is power-law function with $f(r_{ij})= r^{-\alpha}_{ij}  (\alpha > 0)$, and the another one is exponential function with $f(r_{ij}) = e^{-\beta r_{ij}} (\beta > 0)$ \cite{chen2015distance}.  $f(r_{ij})$, $C$ can be chosen as to fit the data with the simple regression.
Form this standard formulation we also define what we term the \textit{flux}, or the expected flow between organizations given their size, as a function of distance,

\begin{equation}
	\label{eq:gravity_flux}
	\widetilde{f_{ij}} = C*m_{i}m_{j}*f(r_{ij})
\end{equation}

\js{I think we need a results, geographical distance with exponential function, emb distance with power-law function. And It will goes to supplement for sure}
We consider separate variants of $f(r_{ij})$ for geographic distance, $g_{ij}$, and the embedding distance, $d_{ij}$, report the best-fit model of each distance.
For geographic distance, we use the power-law function of the gravity law, with $f(g_{ij})= g^{-\alpha}_{ij}$, where $\alpha$ is estimated from the data. 
For embedding distance , we use the exponential function, with $f(d_{ij}) = e^{-\beta d_{ij}}$, where $\beta$ is estimated from the data. 
To estimate parameters $\alpha$ and $\beta$ are estimated by log-transforming each variant of \label{eq:gravity_flux}, obtaining Eq.~\ref{eq:linear_flux_geo} and Eq.~\ref{eq:linear_flux_emb} in the form of linear equations, and fitting parameters using linear regression.

\begin{equation}
	\label{eq:linear_flux_geo}
	\ln(\frac{f_{ij}}{m_im_j}) = \ln(C)  - \alpha \ln(g_{ij})
\end{equation}

\begin{equation}
	\label{eq:linear_flux_emb}
	\ln(\frac{f_{ij}}{m_im_j}) = \ln(C) - \beta d_{ij}
\end{equation}

@@

Let $f_{ij}$ represents the total \textbf{flux}, the number of people who moved between organizations $i$ and $j$. $M_{i}$ and $M_{j}$ represent the size of an organization. According to our gravity model on the embedding space,

\begin{equation}
    \label{eq:gravity_model}
    \frac{f_{ij}}{M_{i} * M_{j}} \propto \exp(\cos(v_i,v_j))
\end{equation}

where the cosine similarity between two vectors, $\cos(v_{i}, v_{j})$, is defined as,

\begin{equation}
    \cos(v_j, v_i) = \frac{v_j \cdot v_i}{|v_j||v_i|} 
\end{equation}

In the case of scientific co-affiliation, all movements are without direction (symmetric), meaning that  $f_{ij} = f_{ji}$.
Let's denote $\mathpzc{N}_i$ as the neighborhood set of organization i, which are organizations with non-zero co-affiliation to organization i.
Then, the total flux to or from $i$ can be written as

\begin{equation}
    \mu_{i} = \sum_{k\in\mathpzc{N}_i} f_{ik} \propto M_i
\end{equation}

with the hypothesis that a similar fraction of the scholars are mobile in each organization. Then, our gravity model can be written with only terms from mobility data,

\begin{equation}
    \label{eq:gravity_model_approx}
    \frac{f_{ij}}{\mu_i * \mu_j} \propto \exp(\cos(v_i,v_j))
\end{equation}

The embedding that best represents the data should closely approximate the true conditional probability defined in
Eq.~\ref{eq:flux_conditional_prob},

\begin{equation}
    \label{eq:flux_conditional_prob}
    p(j|i) = \frac{f_{ij}}{\mu_i} 
\end{equation}

We can then rewrite Eq.~\ref{eq:gravity_model_approx} as,

\begin{equation}
    \label{eq:gravity_relationship}
    \frac{f_{ij}}{\mu_i * \mu_j} = \frac{\mu_i * p(j|i)}{\mu_i * \mu_j} =  \frac{p(j|i)}{\mu_j} \propto \exp(\cos(v_i,v_j))
\end{equation}

Then, the word embedding model will learns vectors under the condition,

\begin{equation}
    \label{eq:gravity_relationship}
      p(j|i) \propto {\mu_j}* \exp(\cos(v_i,v_j))
\end{equation}

which is restricted to, sum of the conditional probability must be one,

\begin{equation}
    \label{eq:gravity_relationship}
    \sum_{j\in\mathpzc{N}_i} p(j|i) = 1
\end{equation}

Then, we can write $p(j|i)$ with $\mu_{j}$

\begin{equation}
    \label{eq:gravity_relationship}
    p(j|i) \propto \frac{{\mu_j}* \exp(\cos(v_i,v_j)}{\sum_{k\in\mathpzc{N}_i} {\mu_k}* \exp(\cos(v_i,v_k))}
\end{equation}

And, if we remove ${\mu_*}$ for simplicity (or for our special case with $\mu_{i} = \mu_{j} = \mu_*$), then $p(j|i)$ can be written as,

\begin{equation}
    \label{eq:real_embedding_relationship}
    p(j|i) \propto \frac{\exp(\cos(v_j, v_i))}{\sum_{k\in\mathpzc{N}_i} \exp(\cos(v_k,v_i))}
\end{equation}

% $p(j|i)$ and ${\mu_j}$ is given from data, and the our model trains $\exp(\cos(v_i,v_j))$ with the data. Then, can we infer the functional form of the $p(j|i)$ which is written with the results of the model?

% Cosine similarity capture the relative importance of neighborhood, because, with same $p(j|i)$, if $\mu_j$ is large, $\exp(\cos(v_i,v_j)$ should small and if $\mu_j$ is small, $\exp(\cos(v_i,v_j)$ should be large. Based on this relationship, cosine similarity capture the relative importance. Simply, we can use $p(j|i)$ as just normalized importance with

% Also, we can validate this proxy with checking inverse relationship, \ref{eq:inverse_relationship}.

% \begin{equation}
%     \frac{f_{ij}}{\mu_i * \mu_j} \propto \frac{\exp(\cos(v_i, v_j))}{\mu_i * \sum_{w\in\mathpzc{N}_j} \exp(\cos(v_w,v_j))} \propto \exp(\cos(v_i,v_j))
% \end{equation}
% \begin{equation}
% \label{eq:inverse_relationship}
%      \mu_i * \sum_{w\in\mathpzc{N}_j} \exp(\cos(v_w,v_j)) \propto C
% \end{equation}

% According to Fig. \ref{fig:inverse_relationship}, $\mu_{i}$ and $\sum_{w \in \mathcal{N}_{j}} \exp(\cos(v_w \cdot v_j))$ have either a negative power-law relationship, or a weak inverse relationship. 
% This relationship may result form hub nodes, or major organizations central to the global network of co-affiliations. 
% A node with large $\mu_{i}$ would be considered a hub node in the network, and following the friendship paradox, neighboring nodes will likely have fewer neighbors than the hub.

% \begin{figure}
%     \centering
%     \includegraphics[width=0.8\columnwidth]{test.png}
%     \caption{Relationship between the conditional probability and the cosine similarity}
%     \label{fig:inverse_relationship}
% \end{figure}

We can check the evidence that Eq.\ref{eq:real_embedding_relationship} is a good approximation functional form of is $p(i|j)$ in Fig. \ref{fig:real_embedding_relationship}, as high values of flow correspond to high Eq.\ref{eq:real_embedding_relationship} values.

%
% Figure - Conditional Probability
%
\begin{figure}
    \centering
    \includegraphics[width=0.8\columnwidth]{\figdir/Math/approxi_proof.png}
    \caption{Relationship between the conditional probability and the cosine similarity}
    \label{fig:real_embedding_relationship}
\end{figure}

This finding relates to the objective function of \textit{word2vec} (skip-gram) model. The skip-gram model attempts to maximize the average log probability of observing a target word in a sentence, $w_{t}$, given the words in the surrounding context window, $w_{t + k}$. This can be written as follows, where $c$ is the size of the context window and $T$ iterates through possible target words:

\begin{equation}
    \label{eq:skipgram_log_prob_condensed}
    \frac{1}{T}\sum_{t=1}^{T} \sum_{-c \le j \le c, j \neq 0} \log p(w_{t+j}|w_t),  
\end{equation}

where the conditional probability of observing one word, $w_{0}$, given another, $w_{1}$, is defined as,

\begin{equation}
    \label{eq:skipgram_conditional_expanded}
    p(w_O|w_I) = \frac{\exp(v'_{w_O} \cdot v_{w_I})}{\sum_{w=1}^{W} \exp(v'_w \cdot v_{w_I})}, 
\end{equation}

where $v_w$ and $v'_w$ are `input' and `output' vector representations of each word in the vocabulary. In the negative sampling strategies, the denominator is not calculated directly. In general, the cosine similarity of in-vectors is used as a measure of affinity. Eq. \ref{eq:skipgram_conditional_expanded} and Eq. \ref{eq:real_embedding_relationship} share a similar functional form (or Eq.~\ref{eq:real_embedding_relationship} is a special case of Eq. ~\ref{eq:skipgram_conditional_expanded}).

The \textit{word2vec} skip-gram model can be an effective model for investigating any kind of mobility trajectory data, stemming from the mathematical relationship between it and the gravity model of mobility.
